A highly-available move operation for replicated trees

### 2019-06-21 paper/move-op.tex
paper: more editing, space saving
More short lines rewritten away.  Down to ~1 paragraph above limit.

Many systems use a tree-structured data model.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in \S\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see \S\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
In summary, our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
Nor do we need to define a remove operation: instead, we can designate a particular tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.

@@

The intuition behind these functions is as follows: when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
The result of this procedure is that the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.

@@

One final type of conflict that we have not discussed so far is multiple child nodes with the same parent and the same metadata.
For example, in a filesystem, two users could concurrently create files with the same name in the same directory.
Our algorithm does not take any measures to prevent such conflict, but simply retains both child nodes.
In practice, the filename collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.

@@

A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
To prove that our algorithm always maintains a tree structure, no matter which operations are applied, we need to prove that these invariants hold.

In fact, we slightly generalise this property: we allow more than one root to exist, so the graph is actually a forest.
This allows an application to move nodes between multiple trees, if desired.
For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.

@@

To prove that \isa{apply\_op} preserves acyclicity, we first show that for any tree produced by \isa{apply\_ops}, there exists a sequence of operations such that applying those operations using \isa{do\_op} invocations alone (not using \isa{undo\_op} or \isa{redo\_op}) results in the same tree.
(This is the case when operations are applied in order of increasing timestamp.)
Then, by the result of the previous paragraph, any tree produced by \isa{apply\_ops} must be acyclic.

@@

Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
Based on the commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
The details of this proof are given in our open source Isabelle proof files~\cite{SourceFiles}.

@@

The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
However, performing an operation on such a system requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.

@@

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
For tree data structures, several approaches have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
\end{itemize}

>>>>

Many systems use a tree-structured data model.
For example:
\begin{itemize}
    \item Most \textbf{filesystems} are organised as a tree with directories as branch nodes and files as leaf nodes.
        % In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems with hardlinks form a DAG; we will focus on trees now and show in \S\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, lists, figures, sections, and so on.
        These can be nested: for example, a paragraph may appear inside a bulleted list item, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Objects may be combined into a \emph{group} and manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may be combined further into higher-level groups, forming a tree.
    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see \S\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
In summary, our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08} and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

The meaning of an operation \isa{Move t p m c} is that at time \isa{t}, the child node \isa{c} is moved to be a child of parent node \isa{p}.
The operation does not specify the old location of \isa{c}; the algorithm simply ``steals'' \isa{c} from wherever it is currently located in the tree, and moves it to \isa{p}.
If \isa{c} does not currently exist in the tree, it is created as a child of \isa{p}.
For this reason we do not need a separate operation type to create tree nodes: simply performing a move with a fresh child node \isa{c} is sufficient to create \isa{c}.
Nor do we need to define a remove operation: instead, we can designate a tree node as the ``trash'', and perform deletion of a node \isa{c} by moving \isa{c} to be a child of the trash node.

@@

Intuitively, when a replica needs to apply an operation with timestamp \isa{t}, it first undoes the effect of any operations with a timestamp greater than \isa{t}, then performs the new operation, and finally re-applies the previously undone operations.
The result of this procedure is that the state of the tree is as if the operations had all been applied in order of increasing timestamp, even though in fact they might have been applied in any order.

@@

One final type of conflict that we have not discussed so far is multiple child nodes with the same parent and the same metadata.
For example, in a filesystem, two users could concurrently create files with the same name in the same directory.
Our algorithm does not prevent such a conflict but simply retains both child nodes.
In practice, the collision would be resolved by making the filenames distinct, e.g.\ by appending a replica identifier to the filenames.

@@

A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
To prove that our algorithm maintains a tree structure, no matter which operations are applied, we prove that these invariants hold.
In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between multiple trees, if desired.
For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.

@@

To prove \isa{apply\_op} preserves acyclicity we show that trees obtained from \isa{apply\_ops} can be built from repeated use of \isa{do\_op} invocations alone, without \isa{undo\_op} or \isa{redo\_op}.
(This is the case when operations are applied in order of increasing timestamp.)
Then, from our previous results, any tree produced by \isa{apply\_ops} is acyclic.

@@

Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs.
Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
We refer the reader to our Isabelle theory files~\cite{SourceFiles}.

@@

The alternative to our optimistic replication approach would be to use a central server (master replica) or consensus protocol that processes all operations in a serialisable order.
Then, performing an operation requires waiting for synchronous communication with a server potentially located on a different continent, taking hundreds of milliseconds (Table~\ref{tab:rpc-times}); by contrast, a local operation taking a fraction of a millisecond is orders of magnitude faster.
Moreover, a local operation requires no waiting for network communication, and thus it supports disconnected operation: mobile devices can read and modify the replicated tree even while disconnected from the Internet, which is not the case with centralised systems.

@@

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
For tree data structures, several approaches have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
\end{itemize}
